\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-05 1-intro.Rmd
modify introduction section

Many data structures have been proposed for spatial (\pkg{sf} by @sf) and temporal (\pkg{tsibble} by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially because analysts usually treat the spatial and temporal dimension of the data separately, without realising the need to create a new data structure. While this approach follows the third tidy data principal [@tidydata] (*Each type of observational unit forms a table*), analysts always need to manually join results from different observational units or combining multiple tables into one for downstream analysis. This additional step doesn't add new operations into the data but can be error prone. \newline

Currently, available spatio-temporal data structure in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses is based on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which has been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal array with the dplyr's data cube structure \pkg{cubelyr} [@cubelyr] as its backend. While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables for each temporal unit; or split them into two separate tables that has the problem of manually joining, mentioned in the previously. None of these packages enjoy both the benefits of being able to separate manipulation in the two dimensions while also keep the data object as a whole. This create a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: the Ireland wind data from \pkg{gstat} is an classic example data that splits variables into spatial (`wind.loc`) and temporal (`wind`) dimension; Bureau of Meteorology (BoM) provides climate observations that are widely applied in agriculture and ecology study; air pollution data. \newline

@@

The rest of the paper will be divided as follows: Section 2 presents the workflow of data manipulation in cubble. Section 3 explains how cubble deals with some advanced considerations  including data with hierarchical structure, data matching, and how cubble fits with existing static and interactive visualisation tools. Section 4 gives examples of the features introduced in the previous two sections with climate and hydrology data. An example on how cubble handle Netcdf data is also provided. Section 5 concludes the paper

>>>>

Spatio-temporal data concerns variables in the spatial coordinates across time. Some variables are invariant to time and can be solely identified by its spatial location while other time varying variables need both the spatial and temporal identifier to identify. Manipulating data in these two levels is usually done separately and the third tidy data principal [@tidydata] recommends each type of observational unit to form a table. Another feature for a spatio-temporal data structure to have is that variables created from subsequent operations can be automatically added to the existing table with the same level. This is better than an output that index the variables with the identifiers since otherwise analysts will need to manually join this output with existing variables in the same level.

<!-- Many data structures have been proposed for spatial (\pkg{sf} by @sf) and temporal (\pkg{tsibble} by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially because analysts usually treat the spatial and temporal dimension of the data separately, without realising the need to create a new data structure. While this approach follows the third tidy data principal [@tidydata] (*Each type of observational unit forms a table*), analysts always need to manually join results from different observational units or combining multiple tables into one for downstream analysis. This additional step doesn't add new operations into the data but can be error prone. \newline -->

Currently, available spatio-temporal data structure in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses is based on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which has been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal array with the dplyr's data cube structure \pkg{cubelyr} [@cubelyr] as its backend. While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables for each temporal unit; or split them into two separate tables that has the problem of manually joining, mentioned in the previously. This creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: the Ireland wind data from \pkg{gstat} is an classic example data that splits variables into its spatial (`wind.loc`) and temporal (`wind`) dimension; Bureau of Meteorology (BoM) provides climate observations widely applied in agriculture and ecology study. \newline

@@

The rest of the paper will be divided as follows: Section 2 presents the workflow of data manipulation in cubble. Section 3 explains how cubble deals with some advanced considerations  including data with hierarchical structure, data matching, and how cubble fits with existing static and interactive visualisation tools. Section 4 gives examples of the features introduced in the previous two sections with climate and hydrology data. An example on how cubble handle Netcdf data is also provided. Section 5 concludes the paper.

### 2022-02-05 main.tex
modify introduction section

Many data structures have been proposed for spatial (\pkg{sf} by
\citet{sf}) and temporal (\pkg{tsibble} by \citet{tsibble}) data in the
R community, while less has been done for spatio-temporal data. The lack
of such tools could potentially because analysts usually treat the
spatial and temporal dimension of the data separately, without realising
the need to create a new data structure. While this approach follows the
third tidy data principal \citep{tidydata} (\emph{Each type of
observational unit forms a table}), analysts always need to manually
join results from different observational units or combining multiple
tables into one for downstream analysis. This additional step doesn't
add new operations into the data but can be error prone. \newline

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. None of
these packages enjoy both the benefits of being able to separate
manipulation in the two dimensions while also keep the data object as a
whole. This create a gap in the software development. The requirement
for such a tool is important given the ubiquity of spatio-temporal
vector data in the wild: the Ireland wind data from \pkg{gstat} is an
classic example data that splits variables into spatial
(\texttt{wind.loc}) and temporal (\texttt{wind}) dimension; Bureau of
Meteorology (BoM) provides climate observations that are widely applied
in agriculture and ecology study; air pollution data. \newline

@@

In the previous example, there has already been some overlapping of the
glyphs for a few stations near (151E, 34S) and (152E, 33S) and this will
be a problem when mapping more stations in the national level.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map. This example shows how to
organise data at both level with \texttt{switch\_key()}. The dataset
used is \texttt{weatherdata::climate\_full}, which records daily
precipitation and maximum/ minimum temperature for 639 stations in
Australia from 2016 to 2020.

A simple kmean algorithm based on the distance matrix is used here to
create 20 clusters. This creates \texttt{station\_nested} as a station
level nested cubble with a cluster column indicating the group each
station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

>>>>

Spatio-temporal data concerns variables in the spatial coordinates
across time. Some variables are invariant to time and can be solely
identified by its spatial location while other time varying variables
need both the spatial and temporal identifier to identify. Manipulating
data in these two levels is usually done separately and the third tidy
data principal \citep{tidydata} recommends each type of observational
unit to form a table. Another feature for a spatio-temporal data
structure to have is that variables created from subsequent operations
can be automatically added to the existing table with the same level.
This is better than an output that index the variables with the
identifiers since otherwise analysts will need to manually join this
output with existing variables in the same level.

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.
\newline

@@

In the previous example, there has already been some overlapping of the
glyphs for a few stations near (151E, 34S) and (152E, 33S) and this will
be a problem when mapping more stations in the national level.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map.

This example shows how to organise data at both level with
\texttt{switch\_key()}. The dataset used is
\texttt{weatherdata::climate\_full}, which records daily precipitation
and maximum/ minimum temperature for 639 stations in Australia from 2016
to 2020. A simple kmean algorithm based on the distance matrix is used
here to create 20 clusters. This creates \texttt{station\_nested} as a
station level nested cubble with a cluster column indicating the group
each station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.
